---
title: ""
output: 
  html_document:
    df_print: paged
    code_folding: hide
    toc: true
    toc_float: true
    toc_collapsed: true
    toc_depth: 3
runtime: shiny
---

# - Introduction

Always in the context of [DataKind's Fall2024 DataKit Challenge #1](https://github.com/datakind/datakit-housing-fall-2024/discussions/1) and moving forward from the [get started](https://github.com/PykaAlexandro/DataKitFall2024/blob/main/DataKitFall2024_challenge1FL_getstarted_shiny.Rmd) analysis, we will here proceed with the take it further part, investigating the renting availability and its impact on the income for a county of choice, with the reasoning that renting it's a faster solution in case of displacements.

```{r, message = FALSE}
#we load the libraries
library(readr)
library(dplyr)
library(ggplot2)
library(DT)
library(tidyr)
library(stringr)
library(shiny)
#we load the data frame
data_1_FL_takeitfurther_part1 <- read_csv("data_1_FL_takeitfurther_part1.csv")
```

# - Assessment of the overall housing stock

We start by enumerating the availability per census tract, as in number of housing units (defined as ["anything from a house to an apartment or even a boat if a person is currently living there"](https://censusreporter.org/topics/housing/)).

NOTE: selecting a different County will change all the graphs and tables in the document to match that selection.

```{r}
#to choose the county
selectInput("county", 
            label = "County",
            choices = factor(sort(unique(data_1_FL_takeitfurther_part1$County))), 
            selected = "Alachua County")
renderPlot({data_1_FL_takeitfurther_part1 %>%
#we pick the county of choice
    filter(County == input$county) %>%
#we build the graph, ordering by the Housing Units count  
    ggplot(aes(reorder(`Census Tract`, desc(`Housing Units Count`)), `Housing Units Count`)) +
    geom_col(fill = "#0072B2") + 
    theme(axis.text.x = element_blank(),
          axis.ticks = element_blank()) +
    labs(x = NULL,
         y = NULL,
         title = "Number of Housing Units per Census Tract")})
```

The graph is ordered by the census tracts with more housing units, to better convey the county average.

<br>

We provide a dynamic table (downloadable as an excel file or a pdf) as well to satisfy the need to pinpoint specific instances:

```{r}
#to choose the county
selectInput("county", 
            label = "County",
            choices = factor(sort(unique(data_1_FL_takeitfurther_part1$County))), 
            selected = "Alachua County")
renderDT({data_1_FL_takeitfurther_part1 %>%
#we pick the county of choice
    filter(County == input$county) %>%
#we select the variables of interest
    select(1, 3) %>%
#we display the counties with most housing units on top
    arrange(desc(`Housing Units Count`)) %>%
#we transform the Census Tract variable to factor, to more easily select its values
    mutate(`Census Tract` = factor(`Census Tract`)) %>%
#we construct a dynamic table
    datatable(filter = "top", 
              rownames = FALSE,
              extensions = 'Buttons',
              options = list(dom = 'tpB',
                             buttons = c('excel', 'pdf')))})
```

<br>

Out of the overall housing units present, not all are occupied, as we show in the following graph:

```{r}
#to choose the county
selectInput("county", 
            label = "County",
            choices = factor(sort(unique(data_1_FL_takeitfurther_part1$County))), 
            selected = "Alachua County")

renderPlot({
#we define a colorblind friendly palette
  cbbPalette <- c("#0072B2", "#D55E00")
  data_1_FL_takeitfurther_part1 %>%
#we pick the county of choice
    filter(County == input$county) %>%
#we transform the data into long format, for easier plotting 
    pivot_longer(cols = c(4, 7), names_to = "Status", values_to = "Number of Housing Units") %>%
#we build the graph, ordering by the number of Housing Units
    ggplot(aes(reorder(`Census Tract`, desc(`Number of Housing Units`)), `Number of Housing Units`, fill = Status)) +
    geom_col() +
    scale_fill_manual(values = cbbPalette) +
    theme(legend.position = "bottom",
          legend.title = element_blank(),
          axis.text.x = element_blank(),
          axis.ticks = element_blank()) +
    labs(x = NULL,
         y = NULL,
         title = "Number of Housing Units per Census Tract, differentiating by Occupancy")})
```

<br>

To individuate specific instances, we can use the following dynamic table:

```{r}
#to choose the county
selectInput("county", 
            label = "County",
            choices = factor(sort(unique(data_1_FL_takeitfurther_part1$County))), 
            selected = "Alachua County")
renderDT({data_1_FL_takeitfurther_part1 %>%
#we pick the county of choice
  filter(County == input$county) %>%
#we calculate the percentage of vacant housing units and we transform census tract into a factor to more easily select its values
  mutate("Vacant %" = round(Vacant / `Housing Units Count` * 100, 2),
         `Census Tract` = factor(`Census Tract`)) %>%
#we select and reorder the variables of interest
  select(1, 3, 4, 7, `Vacant %`) %>%
#we construct a dynamic table
  datatable(filter = "top", 
              rownames = FALSE,
              extensions = 'Buttons',
              options = list(dom = 'tpB',
                             buttons = c('excel', 'pdf')))})
```

<br>

About the occupied housing units, we can differentiate between owned vs rented,

```{r, warning = FALSE}
#to choose the county
selectInput("county", 
            label = "County",
            choices = factor(sort(unique(data_1_FL_takeitfurther_part1$County))), 
            selected = "Alachua County")
renderPlot({
#we define a colorblind friendly palette
  cbbPalette <- c("#0072B2", "#56B4E9")
  data_1_FL_takeitfurther_part1 %>%
#we pick the county of choice
    filter(County == input$county) %>%
#we calculate the percentage of occupied housing units that are rented
    mutate(RentProp = `Renter occupied` / Occupied) %>%
#we transform the data into long format, for easier plotting 
    pivot_longer(cols = 5:6, names_to = "Status", values_to = "Number of Housing Units") %>%
#we build the graph, ordering by RentProp
    ggplot(aes(reorder(`Census Tract`, desc(RentProp)), `Number of Housing Units`, fill = Status)) +
    geom_col(position = position_fill()) +
    scale_fill_manual(values = cbbPalette) +
    theme(legend.position = "bottom",
          legend.title = element_blank(),
          axis.text.x = element_blank(),
          axis.ticks = element_blank()) +
    labs(x = NULL,
         y = NULL,
         title = "Proportions of Owner and Renter Occupied Housing Units per Census Tract")})
```

calculate the overall ownership in percentage,

```{r}
#to choose the county
selectInput("county", 
            label = "County",
            choices = factor(sort(unique(data_1_FL_takeitfurther_part1$County))), 
            selected = "Alachua County")
renderTable({data_1_FL_takeitfurther_part1 %>%
#we pick the county of choice
    filter(County == input$county) %>%
#we calculate the total figures
    summarise("Total Occupied" = sum(Occupied),
              "Total Owner" = sum(`Owner occupied`),
              "Total Renter" = sum(`Renter occupied`)) %>%
#we calculate the percentage of owners
    mutate(`Owner %` = round(`Total Owner` / `Total Occupied` * 100, 2))})
```

<br>

And see what census tracts, if any, have only renters:

```{r}
#to choose the county
selectInput("county", 
            label = "County",
            choices = factor(sort(unique(data_1_FL_takeitfurther_part1$County))), 
            selected = "Alachua County")
renderDT({data_1_FL_takeitfurther_part1 %>%
#we pick the county of choice
    filter(County == input$county) %>%
#we calculate the percentage of rented housing units occupied and we transform census tract into a factor to more easily select its values
    mutate(`Renter %` = round(`Renter occupied` / Occupied * 100, 2),
           `Census Tract` = factor(`Census Tract`)) %>%
#we select the variables of interest
    select(1, 4, 5, 6, `Renter %`) %>%
#we display the census tracts with all housing unites occupied by renters on top
    arrange(desc(`Renter %`)) %>%
#we construct the dynamic table
    datatable(filter = "top", 
              rownames = FALSE,
              extensions = 'Buttons',
              options = list(dom = 'tpB',
                             buttons = c('excel', 'pdf')))})
```

<br>

# - Renting availability and impact on the income.

Among the vacant housing units, here we differentiate between the ones that are available to rent and the ones that aren't, both with a graph:

```{r, warning = FALSE}
#to choose the county
selectInput("county", 
            label = "County",
            choices = factor(sort(unique(data_1_FL_takeitfurther_part1$County))), 
            selected = "Alachua County")
renderPlot({
#we define a colorblind friendly palette
  cbbPalette <- c("#F0E442", "#D55E00")  
  data_1_FL_takeitfurther_part1 %>%
#we pick the county of choice
    filter(County == input$county) %>%
#we calculate the percentage of vacant housing units that are available to rent and we collapse the other instances
    mutate(ToRentProp = `For rent` / Vacant,
           Other = rowSums(pick(9:14))) %>%
#we transform the data into long format, for easier plotting
    pivot_longer(cols = c(8, last_col()), names_to = "Status", values_to = "Number of Housing Units") %>%
#we build the plot, ordering by ToRentProp
    ggplot(aes(reorder(`Census Tract`, desc(ToRentProp)), `Number of Housing Units`, fill = Status)) +
    geom_col(position = position_fill()) +
    scale_fill_manual(values = cbbPalette) +
    theme(legend.position = "bottom",
          legend.title = element_blank(),
          axis.text.x = element_blank(),
          axis.ticks = element_blank()) +
    labs(x = NULL,
         y = NULL,
         title = "Vacancy Status per Census Tract")})
```

<br>

And with a table:

```{r}
#to choose the county
selectInput("county", 
            label = "County",
            choices = factor(sort(unique(data_1_FL_takeitfurther_part1$County))), 
            selected = "Alachua County")

renderDT({data_1_FL_takeitfurther_part1 %>%
#we pick the county of choice
    filter(County == input$county) %>%
#we calculate the percentage of vacant housing units that are available to rent and we transform census tract into a factor to more easily select its values
    mutate(`For Rent %` = round(`For rent` / Vacant * 100, 2),
           `Census Tract` = factor(`Census Tract`)) %>%
#we select the variables of interest
    select(1, 7:14, last_col()) %>%
#we construct the dynamic table
    datatable(filter = "top", 
              rownames = FALSE,
              extensions = 'Buttons',
              options = list(dom = 'tpB',
                             buttons = c('excel', 'pdf')))})
```

<br>

Lastly, we can study how the rents respond to the economic possibilities of the population, starting with the contract rents, defined as [monthly rent agreed to without adjustments for utilities or other payments](https://censusreporter.org/topics/housing/).

```{r}
#to choose the county
selectInput("county", 
            label = "County",
            choices = factor(sort(unique(data_1_FL_takeitfurther_part1$County))), 
            selected = "Alachua County")
renderPlot({data_1_FL_takeitfurther_part1 %>%
#we pick the county of choice
    filter(County == input$county) %>%
#we transform the data into long format, for easier plotting
    pivot_longer(cols = contains("contract"), names_to = "Rent", values_to = "Number") %>%
#we transform the Rent variable for easier visualization
    mutate(Rent = str_sub(Rent, 1, -11),
           Rent = factor(Rent, levels = unique(Rent), ordered = TRUE)) %>%
#we build the graph  
    ggplot(aes(Rent, Number)) +
    geom_col(fill = "#0072B2") +
    coord_flip() +
    labs(x = NULL,
         y = NULL,
         title = "Distribution of Contract Rent")})
```

<br>

As a quick comparison, we present the distribution of the average monthly income per household (calculated from the yearly median one).

```{r, warning = FALSE, message = FALSE}
#we load the income data frame
data_1_FL_getstarted <- read_csv("data_1_FL_getstarted.csv")
#to choose the county
selectInput("county", 
            label = "County",
            choices = factor(sort(unique(data_1_FL_getstarted$county))), 
            selected = "Alachua County")
renderPlot({data_1_FL_getstarted %>%
#we pick the county of choice
    filter(county == input$county) %>%
#we change the negative values (present with missing data) to NAs  
    mutate(across(everything(), ~ case_when(.x >=0 ~ .x))) %>%
#we build the graph  
    ggplot(aes(med_hh_inc_est / 12)) +
    geom_density(fill = "#0072B2") +
    scale_x_continuous(labels = scales::label_dollar()) +
    scale_y_continuous(labels = NULL, breaks = NULL) +
    labs(x = NULL,
         y = NULL,
         title ="Distribution of Average Monthly Income per Census Tract")})
```

<br>

As usual, we provide as well a dynamic table to complete the graph.

```{r}
#to choose the county
selectInput("county", 
            label = "County",
            choices = factor(sort(unique(data_1_FL_takeitfurther_part1$County))), 
            selected = "Alachua County")
renderDT({data_1_FL_takeitfurther_part1 %>%
#we pick the county of choice
    filter(County == input$county) %>%
#we select the variables of interest
    select(1, contains("contract")) %>%
#we transform the Census Tract variable into a factor to easily select between its values
    mutate(`Census Tract` = factor(`Census Tract`)) %>%
#we construct the dynamic table
    datatable(filter = "top",
              rownames = FALSE,
              extensions = c("Buttons", "FixedColumns"),
              options = list(dom = 'tpB',
                             buttons = c('excel', 'pdf'),
                             scrollX = TRUE, 
                             fixedColumns = list(leftColumns = 1)))})
```

We can perform the same analysis for gross rent, defined as contract rent plus the average cost of the utilities (electricity, gas, and water and sewer) and fuel (oil, coal, kerosene, wood, etc), noting an obvious shift to more expensive rents.

```{r, message = FALSE}
#to choose the county
selectInput("county", 
            label = "County",
            choices = factor(sort(unique(data_1_FL_takeitfurther_part1$County))), 
            selected = "Alachua County")
renderPlot({data_1_FL_takeitfurther_part1 %>%
#we pick the county of choice
    filter(County == input$county) %>%
#we transform the data into long format, for easier plotting   
    pivot_longer(cols = contains("gross"), names_to = "Rent", values_to = "Number") %>%
#we transform the Rent variable for an easier visualization
    mutate(Rent = str_sub(Rent, 1, -8),
           Rent = factor(Rent, levels = unique(Rent), ordered = TRUE)) %>%
#we build the graph    
    ggplot(aes(Rent, Number)) +
    geom_col(fill = "#0072B2") +
    coord_flip() +
    labs(x = NULL,
         y = NULL,
         title = "Distribution of Gross Rent")})
renderDT({data_1_FL_takeitfurther_part1 %>%
#we pick the county of choice
    filter(County == input$county) %>%
#we select the variable of interest
    select(1, contains("gross")) %>%
#we transform the Census Tract variable into a factor to easily select between its values
    mutate(`Census Tract` = factor(`Census Tract`)) %>%
#we construct the dynamic table
    datatable(filter = "top",
              rownames = FALSE,
              extensions = c("Buttons", "FixedColumns"),
              options = list(dom = 'tpB',
                             buttons = c('excel', 'pdf'),
                             scrollX = TRUE, 
                             fixedColumns = list(leftColumns = 1)))})
```

<br>

Census provides us also with the percentage of income spent on gross rent.

```{r, message = FALSE}
#to choose the county
selectInput("county", 
            label = "County",
            choices = factor(sort(unique(data_1_FL_takeitfurther_part1$County))), 
            selected = "Alachua County")
renderPlot({data_1_FL_takeitfurther_part1 %>%
#we pick the county of choice
    filter(County == input$county) %>%
#we transform the data into long format, for easier plotting  
    pivot_longer(cols = 65:last_col(), names_to = "RentPerc", values_to = "Number") %>%
#we transform the Census Tract variable into a factor to better visualize it on the graph
    mutate(Rent = factor(RentPerc, levels = unique(RentPerc), ordered = TRUE)) %>%
#we build the graph  
    ggplot(aes(RentPerc, Number)) +
    geom_col(fill = "#0072B2") +
    coord_flip() +
    labs(x = NULL,
         y = NULL,
         title = "Distribution of Percentage of Income spent on Gross Rent")})
renderDT({data_1_FL_takeitfurther_part1 %>%
#we pick the county of choice
    filter(County == input$county) %>%
#we select the variables of interest
    select(1, 65:last_col()) %>%
#we transform the Census Tract variable into a factor to easily select between its values
    mutate(`Census Tract` = factor(`Census Tract`)) %>%
#we construct the dynamic table
    datatable(filter = "top",
              rownames = FALSE,
              extensions = c("Buttons", "FixedColumns"),
              options = list(dom = 'tpB',
                             buttons = c('excel', 'pdf'),
                             scrollX = TRUE, 
                             fixedColumns = list(leftColumns = 1)))})
```
